Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-26 paper.tex
ENH:  BYU vs. Boise State.

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}, can be viewed as an alternative smoothing approach in which explicit regularization is achieved through fast B-spline approximation.
This characterization can be used to provide alternative B-spline ``flavored'' diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open labeled brain data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological
sciences.  Neuroscience research efforts, such as characterizing
brain morphology, require accurate and robust methods for
producing such mappings.  The
extensive literature detailing methodology is evidence of the rich history of
algorithmic development which continues contemporaneously.
We highlight several key historical contributions which are particularly relevant to the work presented.

Free-form deformation (FFD) image registration, characterized by  regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity,
good performance, desirable properties, and several available
implementations.  Current research was
preceded by related work for
geometric modeling \citep{sederberg1986} and originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.
Continued development within this early spline-based paradigm produced additional innovations such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation constraints \citep[e.g.][]{rohlfing2003}, and notable open source implementations \citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

@@

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration literature.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric.
By explicitly symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN) minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}
\end{align}
with extension to arbitrary similarity metric choice, i.e. the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

@@

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common, contemporary processing infrastructures.

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is possible since the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  As mentioned earlier, 
a viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient indexed by $c$.

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass. The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Note that any pair of images can be specified per stage.

Although the
resulting transforms for each stage can be written to disk as output, the 
default output consists of a condensed set of transforms where compatible 
transforms have been composed to a single transform.  For example, in the
above command call, the initial translation, rigid, and affine transforms are
combined into a single generic affine transform file with the results of the
deformable transform consisting of discrete vector fields.  The output
transform files can then be applied using the {\tt antsApplyTransforms}
program which permits composition of any number of transform files with 
various interpolation schemes.   For both programs interpolation is never
performed more than once.

@@

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  This was accomplished using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

@@

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with the ANTs repository.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus
labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.

@@

\begin{figure}[htb]
  \centering
  \begin{tabular}{c}
  \includegraphics[width=0.95\textwidth]{evaluation.jpg}
  \end{tabular}
  \caption{Dice results for both algorithms for 
  each subject warped to every other subject using the pseudo-
  geodesic transform.  Each 
  row corresponds to one of the five data sets used for evaluation.
  For each data set we include a plotting of all individual label 
  Dice results by volume and a combined label box plotting.  The
  left and right halves show the results for MALF-derived and 
  original labelings, respectively.
  }
  \label{fig:dice}
\end{figure}

@@

As mentioned previously, a template was constructed for each data
set (cf Figure \ref{fig:templates}) from all cohort images.  Subsequently, 
each image was registered to its corresponding template using either 
SyN or B-spline SyN as described previously (prior linear registration 
stages were identical between the two algorithms).  As a brute-force parameter
exploration is not a part of this work, we rely on previously
reported research \citep{klein2009,avants2011} and our own experience as
authors/developers of the algorithm/software to select parameters which 
demonstrate robust performance across data sets.  For both algorithms,
the gradient step was 0.1 during each of the four multi-resolution levels
with shrink factors of $\{6,4,2,1\}$ with Gaussian smoothing for each
of those levels being $\mathcal{N}\left(0,\{9,4,1,0\}\right)$ in terms
of voxels.  The number of iterations per level were $\{100,100,70,20\}$
with a convergence threshold of $10^{-9}$ and window size of 15 iterations.

All processing was performed using the computational cluster at the 
University of Virginia.%
\footnote{
http://www.uvacse.virginia.edu
}
All perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
It should be noted that for the
data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.

The differences between the two algorithms consist of the Gaussian and 
B-spline parameters governing the update field smoothing, $S_v$.  In our 
experience, smoothing of the total field did not improve the results, at
least for these data (which conforms with our experience with other
data), so the total field smoothing, $S_{\phi}$ is 0 for
both registration approaches.  Specifically, the chosen parameters
for the SyN algorithm were:  $S_{\phi} = \mathcal{N}(0,0)$ and 
$S_{v} = \mathcal{N}(0,3)$ in voxel terms.%
\footnote{
 Or, in equivalent {\tt antsRegistration} command line parlance, {\tt -t
 SyN[0.1,3,0]}.
 }  Although our experience with B-spline SyN is much more limited, we 
were able to choose comparable parameters based on a knot spacing
for the update field of 26 mm at the base level which is reduced by
a factor of two for each subsequent multiresolution level.  This 
yields a final knot spacing of 3.25 mm.%
\footnote{
 Or, equivalently, {\tt -t
 BSplineSyN[0.1,26,0]}.
% The {\tt antsRegistration} command line currently 
% permits specification of the
% B-spline total and update field smoothing only in terms of total mesh size over
% the entire image domain, i.e. {\tt -t BSplineSyN[0.1,10x10x8,0]}.  However, 
% in order to minimize anisotropic smoothing, we determined for each subject 
% the mesh size closest to producing a knot spacing of 26 mm which was typically
% around 10 mesh elements per dimension.  We are currently planning on modifying
% the command line to permit specification of a knot size (similar to the 
% ANTs-based N4 bias correction algorithm \citep{tustison2010} which also uses the
% same underlying B-spline computational machinery).
 }
As specified in the supplementary material associated with \cite{klein2009}, 
this is similar to the 
gradient smoothing parameter for the IRTK FFD algorithm
which also used four multi-resolution levels with an initial knot spacing of
20 mm per dimension for a final knot spacing of 2.5 mm.

Quality of overlap  using the 
Dice similarity metric 
was determined from the transformed labels
using the open source ITK
implementation described in \cite{tustison2009a}.
Both the original labels and MALF labels were warped
to the fixed image for comparison using nearest neighbor
interpolation.  A joint Dice metric value was calculated
from the combined labels for each cohort for each of the
two SyN methods.  These values are rendered in notched
box plot format in Figure \ref{fig:dice}.  Non-overlapping notches
indicate approximately statistically significantly different median values
at the 95\% confidence level \citep{mcgill1978}.
For all data sets, the B-spline SyN variant showed a small but statistically
significant improvement in overall Dice values.
In order to provide a more complete picture of performance differences,
we also accounted for label volumetric considerations \citep{rohlfing2012}.  In Figure
\ref{fig:dice} we plotted the Dice value difference between each
SyN variant (B-spline SyN - SyN) for each label of each intra-subject
registration pair within a data set versus the volume of the label
in the fixed image.

@@

To further characterize the deformable transform differences, we 
calculated the log of the Jacobian determinant of the transformations from 
each subject to the template and 
tabulated statistical information within the brain region only. 
A noticeable difference between the two algorithms was the
respective range of values in log Jacobians.  We plotted the 
(95th\% - the 5th\%) for each algorithm across all data sets
 in Figure \ref{fig:jacobian}.  It is apparent 
that B-spline SyN produces a much greater range of deformation
values during the course of optimization.

@@

A significant amount of research has been devoted to image registration
algorithmic development.  Given their many salient characteristics particularly with
respect to large deformation estimation constrained by topological 
continuity, diffeomorphic registration
approaches have been a particular focus in the neuroimaging community.
However, many groups continue to find success with non-diffeomorphic
FFD methods
\citep[e.g.][]{rueckert1999,klein2010}.  Using our DMFFD framework, 
B-spline regularization is easily adapted into the diffeomorphic registration 
framework and performs comparably to analogous algorithms which we demonstrated 
in this work for the case of the widely-used SyN.

B-spline SyN produced slightly greater Dice values than the original SyN.
Interestingly it was shown that the range of log Jacobian values tend to be
significantly higher for the former than the latter.  Although possible 
explanations include the approximation-versus-convolution distinction mentioned
earlier, further exploration is planned.

One of the advantages that has not been explored in this manuscript is 
the use of B-spline SyN for small deformation estimation problems such 
as in pulmonary or cardiac applications.  Such problems typically require
greater regularization which implies larger discrete Gaussian kernels.
A related issue is applications involving severely anisotropic data where
the continuous nature of the DMFFD approach might help over Gaussian convolution.
Additionally, we have not yet explored the use of sparsely sampled data 
and the advantages of the proposed framework.
Ongoing work will continue to explore these issues.

>>>>

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, the {\it directly manipulated free-form deformation} approach \citep{tustison2009}, can be viewed as a smoothing alternative in which explicit regularization is achieved through fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image registration solutions with several advantages.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open labeled brain data and open source evaluation tools.  
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological
sciences.  Neuroscience research efforts, such as characterizing
brain morphology, require accurate and robust methods for
producing such mappings.  The
extensive literature detailing methodology is evidence of the rich history of
algorithmic development which continues contemporaneously.
We highlight several key historical contributions which are particularly relevant to the work 
subsequently presented.

Free-form deformation (FFD) image registration, characterized by  regularization based on the B-spline basis functions, has several
advantages including algorithmic simplicity,
good performance, desirable properties, and several available
implementations.  Current research was
preceded by related work for
geometric modeling \citep{sederberg1986} and originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.
Continued development within this early spline-based paradigm produced additional innovations such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation constraints \citep[e.g.][]{rohlfing2003}, and notable open source implementations \citep[e.g.][]{ibanez2005,klein2010,modat2010,shackleford2010}.

@@

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration community.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric.
By explicitly symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN) minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same parameterized path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}.
\end{align}
With extension to arbitrary similarity metric choice, the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

@@

Although presenting a rigorous framework for image registration solutions
with desirable properties, the complexity of these diffeomorphic methodologies
requires substantial computational resources.  For typical 3-D neuroimaging
applications, the corresponding solutions require numerical integration over
and storage of 4-D velocity fields at each iteration which is limiting for
many common computational platforms.

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is due to the restriction of the discrete, time-parameterized
velocity field samples to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  As mentioned earlier, 
a viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation, spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient at voxel $c$.

@@

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0GenericAffine.mat \
                    --default-value 0
\end{lstlisting}
In this example, we first calculate an initial translation transform by aligning
the centers of (intensity) mass (although alignment based on other features is possible). 
The resulting transform is then used as input 
for determining an optimal rigid transform.  Serial propagation of the resulting
composite transform continues until all optimal transforms have been determined.
Optimization for each stage is determined by the specified general parameters 
including: smoothing and downsampling of fixed and moving images, convergence
criteria (including number of iterations per resolution level) and metric (or
metrics).  Note that any pair of images can be specified per metric per stage.

Although the
resulting transforms for each stage can be written to disk as output, the 
default output consists of a condensed set of transforms where compatible 
transforms have been composed to a single transform.  For example, in the
above command call, the initial translation, rigid, and affine transforms are
combined into a single generic affine transform file with the results of the
deformable transform consisting of discrete vector fields.  The output
transform files can then be applied using the {\tt antsApplyTransforms}
program which permits composition of any number of transform files with 
different interpolation schemes.   For both programs interpolation is never
performed more than once.

@@

Comparative evaluation of the two SyN registration approaches was performed 
within each cohort using a ``pseudo-geodesic'' approach.  Instead of
registering every subject to every other subject within a data set,
we generated the transforms from each subject to a cohort-specific
shape/intensity template.  These templates were built using the ANTs script {\tt antsMultivariateTemplateConstruction.sh}
which is a multivariate implementation of the work described in
\cite{avants2010}. Canonical views for each of the five templates
used for this study are given in Figure \ref{fig:templates}.  
Since calculation of the transform from each subject to the template also
includes generation of the corresponding inverse transform, the total
transformation from a given subject to any other is determined from the composition
of transforms mapping through the template.  An example illustration of the
geodesic approach is given in Figure \ref{fig:pseudo_geodesic}.

@@

Additionally, we refined the labelings for each subject of each cohort
using the multi-atlas label fusion algorithm (MALF) developed by \cite{wang2012}
which is also distributed with ANTs.
For a given subject within a data set, every other subject was mapped to 
that subject using the pseudo-geodesic transform.  The set of transformed labelings 
were then used to determine a consensus
labeling for that subject.
This was to minimize the obvious {\it observer dimensionality artifacts} where manual
raters observe and label in a single dimension at a time.  This
is most easily seen in the axial or sagittal views of the different cohorts 
as labelings were done primarily in the coronal view (see Figure \ref{fig:labels}).  
We include both sets of results.  This provides two sets of labels per subject
for evaluation.

@@

\begin{figure}[htb]
  \centering
  \begin{tabular}{c}
  \includegraphics[width=0.95\textwidth]{evaluation.jpg}
  \end{tabular}
  \caption{Dice results for both algorithms for 
  each subject warped to every other subject using the pseudo-
  geodesic transform.  Each 
  row corresponds to one of the five data sets used for evaluation.
  For each data set we include a plotting of all individual label 
  Dice results by volume and a combined label box plotting.  The
  left and right halves show the respective results for 
  original and MALF-derived labelings.
  }
  \label{fig:dice}
\end{figure}

@@

As mentioned previously, a template was constructed for each data
set (cf Figure \ref{fig:templates}) from all cohort images.  Subsequently, 
each image was registered to its corresponding template using either 
SyN or B-spline SyN as described previously (prior linear registration 
stages were identical between the two algorithms).  As a brute-force parameter
exploration is not a part of this work, we rely on previously
reported research \citep{klein2009,avants2011} and our own experience as
authors/developers of the algorithm/software to select parameters which 
demonstrate robust performance across data sets.  For both algorithms,
the gradient step was 0.1 for each of the four multi-resolution levels
with shrink factors of $\{6,4,2,1\}$ and Gaussian smoothing for each
of those levels being $\mathcal{N}\left(0,\{9,4,1,0\}\right)$ in terms
of voxels.  The number of iterations per level were $\{100,100,70,20\}$
with a convergence threshold of $10^{-9}$ and window size of 15 iterations.

All processing was performed using the computational cluster at the 
University of Virginia.%
\footnote{
http://www.uvacse.virginia.edu
}
The perl scripts used to create the jobs for the cluster are
included in the github account associated with this evaluation.  
It should be noted that for the
data sets used in this study, times for B-spline SyN were approximately 
15-40\% greater than Gaussian-based SyN using single-threading and a dense
metric gradient sampling.

The differences between the two algorithms consist of the Gaussian and 
B-spline parameters governing the update field smoothing, $S_v$.  In our 
experience, smoothing of the total field did not improve the results, at
least for these data (which conforms with our experience with other
data), so the total field smoothing, $S_{\phi}$ is 0 for
both registration approaches.  Specifically, the chosen parameters
for the SyN algorithm were:  $S_{\phi} = \mathcal{N}(0,0)$ and 
$S_{v} = \mathcal{N}(0,3)$ in voxel terms.%
\footnote{
 Or, in equivalent {\tt antsRegistration} command line parlance, {\tt -t
 SyN[0.1,3,0]}.
 }  Although our experience with B-spline SyN is much more limited, we 
were able to choose comparable parameters based on a knot spacing
for the update field of 26 mm at the base level which is reduced by
a factor of two for each succeeding multiresolution level.  This 
yields a final knot spacing of 3.25 mm.%
\footnote{
 Or, equivalently, {\tt -t
 BSplineSyN[0.1,26,0]}.
% The {\tt antsRegistration} command line currently 
% permits specification of the
% B-spline total and update field smoothing only in terms of total mesh size over
% the entire image domain, i.e. {\tt -t BSplineSyN[0.1,10x10x8,0]}.  However, 
% in order to minimize anisotropic smoothing, we determined for each subject 
% the mesh size closest to producing a knot spacing of 26 mm which was typically
% around 10 mesh elements per dimension.  We are currently planning on modifying
% the command line to permit specification of a knot size (similar to the 
% ANTs-based N4 bias correction algorithm \citep{tustison2010} which also uses the
% same underlying B-spline computational machinery).
 }
For comparison, after selecting these smoothing parameters we discovered 
in the supplementary material of \cite{klein2009} the similarity to the 
gradient smoothing parameter for the IRTK FFD algorithm
which also used four multi-resolution levels with an initial knot spacing of
20 mm per dimension for a final knot spacing of 2.5 mm.

Quality of overlap  using the 
Dice similarity metric 
was determined from the transformed labels
using the open source ITK
implementation described in \cite{tustison2009a}.
Both the original labels and MALF labels were warped
to the fixed image for comparison using nearest neighbor
interpolation.  A joint Dice metric value was calculated
from the combined labels for each cohort for each of the
two SyN methods.  These values are rendered in notched
box plot format in Figure \ref{fig:dice}.  Non-overlapping notches
indicate approximately statistically significantly different median values
at the 95\% confidence level \citep{mcgill1978}.
For all data sets, the B-spline SyN variant showed a small but statistically
significant improvement in overall Dice values.
In order to provide a more complete picture of performance differences,
we also accounted for label volumetric considerations \citep{rohlfing2012}.  In Figure
\ref{fig:dice} we plotted the Dice value difference between each
SyN variant (B-spline SyN - SyN) for each label of each intra-subject
registration pair within a data set versus the volume of the label
in the fixed image.  Values above and below the dashed line indicate better 
regional performance for B-spline SyN and SyN, respectively.

@@

To further characterize the deformable transform differences, we 
calculated the log of the Jacobian determinant of the transformations from 
each subject to the template and 
tabulated statistical information within the brain region only. 
A noticeable difference between the two algorithms was the
respective range of values in log Jacobians.  We plotted the 
(95th\% - 5th\%) for each algorithm across all data sets
 in Figure \ref{fig:jacobian}.  It is apparent 
that B-spline SyN results exhibit a much greater range of deformation.

@@

B-spline SyN produced slightly greater Dice values than the original SyN.  Although
actual differences are relatively small, they are statistically significant. By 
implementing both algorithms in the the same code base, we are not only able to 
eliminate non-regularization components of the registration but we are also able
to eliminate implementation differences.  Thus, performance disparity can be isolated
to smoothing choice.  However, even with this restricted focus there are various
reasons for the evaluation outcome.  These include the approximation-versus-convolution 
distinction mentioned earlier for the two regularization approaches (which could
also explain the reason why the range of log Jacobian values tend to be
significantly higher for B-spline SyN).  Also, the fact that the regularization 
for B-spline SyN is theoretically continuous whereas the truncated Gaussian 
convolution is only a discrete approximation could be a potential factor.  Exploration
of other possibilities are planned for future work.

One of the advantages that has not been explored in this manuscript is 
the use of B-spline SyN for small deformation estimation problems such 
as in pulmonary or cardiac applications.  Such problems typically require
greater regularization which implies larger discrete kernels for 
Gaussian convolution.
A related issue is applications involving severely anisotropic data where
the continuous nature of the DMFFD approach might help over Gaussian convolution.
Additionally, we have not yet explored the use of sparsely sampled data 
and the advantages of the proposed framework.
Ongoing work will continue to explore these issues.

A significant amount of research has been devoted to image registration
algorithmic development.  Given their many salient characteristics particularly with
respect to large deformation estimation constrained by topological 
continuity, diffeomorphic registration
approaches have been a particular focus in the neuroimaging community.
However, many groups continue to find success with non-diffeomorphic
FFD methods
\citep[e.g.][]{rueckert1999,klein2010}.  Using our DMFFD framework, 
B-spline regularization is easily adapted into the diffeomorphic registration 
framework and performs well compared to analogous algorithms which we demonstrated 
in this work for the case of the widely-used SyN.
